---
title: "Calculating Fish Metrics"
author: "Karen Blocksom "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Fish Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1 Preliminaries

One of the primary goals of this package is to calculate fish metrics as used in the National Rivers and Streams Assessment (NRSA) conducted in 2008-9 and 2013-14 by EPA's Office of Water. Fish were collected at all sites where present, and the common name, number of individuals, and number of anomalies were recorded in the field. Non-native status was then determined for all taxa on a site-specific basis. For the purposes of calculations, the metrics were broken down into several subsets, depending on which autecology traits were required to calculate them. In addition, if the non-native status is included with input data, the native version of metrics is also calculated, as well as alien species metrics, if desired. There are also separate functions to calculate just the metrics used in the regional multimetric indices (MMIs) and the MMI scores themselves based on metric inputs. This vignette will describe how to use the various functions along with the example datasets provided and the included taxa lists, to obtain the full set of metrics, a subset of metrics based on metric type, and the fish multimetric index used in assigning biological condition.

The functions in this package allow for a great deal of flexibility in terms of input variable names and the taxalist used in calculating metrics. However, the actual values of various traits, and the fish names as used in the taxalist are assumed to match those used in NRSA. This means that some metrics will not calculate correctly if the NRSA taxalist is not used. In addition, the functions assume that data have already been summarized to the level of sample and taxon, and a within-sample distinctness (0/1) already assigned to each taxon, so any steps required to get to this point must be done prior to use of the functions. All taxa in the count data must be included in the taxa list in order to be included in the calculations. Functions will look for specific types of inconsistencies, like missing variables, and either provide a warning or will fail to run, depending on the issue.  

********
# 2 Calculating NRSA fish metrics

There are two main approaches to calculating the fish metrics. One may either calculate all fish metrics at once, which may take a long time with larger datasets, or one may calculate various subsets of metrics. The latter approach may be helpful if only certain metrics are of interest or only certain traits are available. In the first approach, the function `calcAllFishMets` serves as a wrapper function for a number of metric subset functions. The other approach involves using individual funcions, including `calcFishTaxMets`, `calcFishTolMets`, `calcFishTrophicMets`, `calcFishOtherMets`, `calcFishNativeMets`, and/or `calcFishAnomMets`. The metrics used in regional fish MMIs and the MMIs themselves, can be calculated using the functions `calcNRSA_FishMMImets()` and `calcFishMMI()`, respectively.

## 2.1 All metrics at once

Calculate all metrics using the example fish count data included in the package, `fishEx`, along with the included taxalist, `fishTaxa`.

```{r allFish.1}
library(aquamet)

head(fishEx)
```
This example dataset contains all of the necessary variables required to calculate all of the fish metrics. For this dataset, `UID` identifies individual samples, `TAXA_ID` identifies the taxon and links to same field in the taxalist, `FINAL_CT` is the count for the taxon, `ANOM_CT` is the number of anomalies for a given taxon, `IS_DISTINCT` is the distinctness, which 0 indicating non-distinct within the sample and 1 distinct, and `NON_NATIVE` indicates whether the taxon is non-native in this particular site (Y=YES, N=NO).

```{r allFish.2}
head(fishTaxa)
```
This is the desired format of the taxalist. All of the necessary traits are included in the taxalist for metric calculation. The variable `FAM_OR_CLS` is the same as `FAMILY` and is an artifact from the NRSA 2008-9 study.
```{r allFish.3}
outdf <- calcAllFishMets(fishEx,fishTaxa,sampID=c("UID"), dist="IS_DISTINCT",
                            ct="FINAL_CT",anomct='ANOM_CT',taxa_id='TAXA_ID',tol='TOLERANCE_NRSA'
                            , tolval='TOL_VAL_EMAPW', vel='VEL_NRSA', habitat='HABITAT_NRSA'
                            , trophic='TROPHIC_NRSA', migr='MIGR_NRSA', nonnat='NON_NATIVE'
                            , reprod='REPROD_NRSA', temp='TEMP_NRSA', family='FAMILY', genus='GENUS'
                            , comname='FINAL_NAME')

names(outdf)
```
The definitions of all metrics can be found in **NRSA_Fish_Metric_Descriptions.pdf** found in this package.

## 2.2 Subsets of metrics

Suppose only tolerance-related and taxonomic metrics are of interest. Then we need only to run the `calcFishTolMets()` and `calcFishTaxMets()` functions. For each of these functions, if we include a variable for `nonnat` in the input data frame, native status versions of metrics will also be calculated. For the purposes of demonstration, we will included this non-native status variable for the tolerance metrics but not the taxonomy metrics.

### Taxonomy metrics

```{r subMets.1}
outTax <- calcFishTaxMets(fishEx, fishTaxa, sampID='UID', dist='IS_DISTINCT',
                          ct='FINAL_CT', taxa_id='TAXA_ID', family='FAMILY', genus='GENUS',
                          comname='FINAL_NAME')
names(outTax)
head(outTax)
```

### Tolerance metrics

For the tolerance metrics, we will suppose we only have trait information for or care about tolerance and habitat, so only metrics based on tolerance alone and tolerance in combination with habitat will be calculated. Since we will also specify the name of the non-native status variable in the input dataset, native versions of metrics will also be calculated.

```{r subMets.2}
outTol <- calcFishTolMets(fishEx, fishTaxa, sampID='UID', dist='IS_DISTINCT',
                          ct='FINAL_CT', taxa_id='TAXA_ID', tol='TOLERANCE_NRSA',
                          tolval='TOL_VAL_EMAPW', nonnat='NON_NATIVE', 
                          habitat='HABITAT_NRSA')
names(outTol)
head(outTol)

```
**************************
# 3 Calculate only fish MMI metrics and MMI

First, we will need to create a data frame containing the three-letter code for the 9 aggregated ecoregions and add that to the input data frame (`indata`). Then run the function `calcNRSA_FishMMImets()`, followed by `calcFishMMI()`.

```{r mmiMets.1}
eco9 <- data.frame(UID=c(10000,10001), ECO9=c('CPL','WMT'),stringsAsFactors=F)

fishEx.1 <- merge(fishEx, eco9, by='UID')
```

## 3.1 Calculate fish MMI metrics

This function calculates metrics for only the MMI for the ecoregion of each site.
```{r mmiMets.2}
outMets <- calcNRSA_FishMMImets(fishEx.1,fishTaxa, "UID", ecoreg='ECO9'
                                 ,dist="IS_DISTINCT",ct="FINAL_CT"
                                 ,taxa_id='TAXA_ID',tol='TOLERANCE_NRSA'
                                 ,vel='VEL_NRSA', habitat='HABITAT_NRSA'
                                 ,trophic='TROPHIC_NRSA', migr='MIGR_NRSA', nonnat='NON_NATIVE'
                                 ,reprod='REPROD_NRSA', family='FAMILY', genus='GENUS'
                                 ,comname='FINAL_NAME')

head(outMets)

```
Note that there are missing values for metrics that are not included in the MMI for a given ecoregion. Any given UID should only have values for 8 metrics, which is the case in our example. 

## 3.2 Calculate NRSA fish MMI

Now use the output of the function above to calculate the fish MMI. To calculate the fish MMI, we have to supply the log10 of watershed area because some metrics in each MMI are adjusted for watershed size. Using the approach above, we create a data frame with this information and merge it with the `outMets` data frame.

```{r mmiMets.3}
wsarea <- data.frame(UID=c(10000,10001), LWSAREA=c(2.1, 3.5), stringsAsFactors=F)

outMets.1 <- merge(outMets, wsarea, by='UID')

outMMI <- calcFishMMI(outMets.1, sampID='UID', ecoreg='ECO9', lwsarea='LWSAREA')

head(outMMI)

```
Again, note that there are missing values for the metrics and metric scores not included in the MMI for a given site. If there is a mix of ecoregions in the input dataset, there will be these missing values for some metrics and metric scores.

# 4 Assign NRSA fish condition

Using the output from \code{calcFishMMI()}, we can now assign fish assemblage condition to these sites using the
\code{assignFishCondition()} function. Additional variables, including watershed area (km2) and total number of individuals 
collected would be required to assign condition only if we had missing MMI values. Since we do not, we do not need to add
these variables. See the example for \code{assignFishCondition} to see examples for that scenario.

```{r fish condition}
outCond <- assignFishCondition(outMMI, sampID='UID', ecoreg='ECO9', mmi='MMI_FISH')

outCond

```
